Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus

### 2024-11-14 paper/paper.tex
tweak

\paragraph{Weak and Strong Equivalence}
In the theory of formal grammars, there are two different notions of
equivalence: weak generative capacity, meaning just which strings are
accepted by the grammar, and \emph{strong} generative capacity, when
the parse trees of the two grammars are isomorphic
\cite{chom1963}. Using linear types as grammars, we can define both of
these notions of equivalence in \theoryabbv.

\begin{definition}
  \label{def:weakequiv}
  Grammars $\A$ and $\B$ are \emph{weakly equivalent} if there exists
  parse transformers $\f : \ltonl{(\A \lto \B)}$ and $\g : \ltonl {(\B
    \lto \A)}$. $\A$ is a \emph{retract} of $\B$ if $\A$ and $\B$ are
  weakly equivalent and $\lambda a. g(f(a)) \equiv \lambda a.a$. They
  are \emph{strongly equivalent} if further the other composition is
  the identity, i.e., $\lambda b. f(g(b)) \equiv \lambda b.b$.
\end{definition}

That is, grammars are weakly equivalent if we can construct a parse
tree of $B$ from a parse tree of $A$ and vice-versa, whereas they are
strongly equivalent if these processes are inverse to each other. The
definition of retract is a useful intermediate notion. In retrospect,
our axiom about $\String$ says that $\String$ is a retract of $\top$.

@@

\paragraph{Ambiguity}
A formal grammar $\A$ is ambiguous if there are multiple parse trees
for the \emph{same} string $\w$.  For example, $\a \oplus \a$ is
ambiguous because there are two parses of $\stringquote{a}$,
constructed using $\inl$ and $\inr$. On the other hand, a formal
grammar is unambiguous when there is at most one parse tree for any
input string. We can capture this notion as a type in \theoryabbv in a
clever way:
\begin{definition}
  \label{def:unambig}
  A grammar $\A$ is \emph{unambiguous} if for every linear type $\B$,
  $\f : \ltonl{(\B \lto \A)}$, and $\g : \ltonl{(\B \lto \A)}$
  then $\f \equiv \g$
\end{definition}
\cref{def:unambig} can be read more intuitively as stating that $\A$
is unambiguous if there is at most one way to transform parses of any
other grammar $\B$ into parse of $\A$. This notion of an unambiguous
type is the analog for linear types of the definition of a (homotopy)
\emph{proposition} in the terminology of homotopy type
theory\cite{hottbook}. The most basic unambiguous types are $\top$ and
$0$, and in a system of classical logic all unambiguous types would
have to be equivalent to one of these, but with our axioms we can show
also that $\I$ and literals $\literal c$ are unambiguous. To see this, first, we
establish two useful properties of unambiguity.
\begin{lemma}[\agdalogo]
  If $\B$ is unambiguous and $\A$ is a retract of $\B$ then $\A$ is unambiguous.
  If a disjunction $\bigoplus\limits_{x:X} A(x)$ is unambiguous then
  each $A(x)$ is unambiguous.
\end{lemma}
From the first principle, we can prove that $\String$ is unambiguous,
since it is a retract of $\top$. In fact, observe that if $\A$ is a
retract of $\B$ and $\B$ is unambiguous, then in fact $\A$ and $\B$
are strongly equivalent, as the equation $\lambda b. f(g(b)) \equiv
\lambda b. b$ follows because $\B$ is unambiguous. Therefore
$\String$ is also strongly equivalent to $\top$. Next, since $\String$ is
defined as a Kleene star, we can easily show that $\String \cong \I
\oplus \Char \oplus (\Char \otimes \Char \otimes \String)$. Then by
the second principle we have that $\I$ and $\Char$ and each literal
$\literal c$ are unambiguous as well.

@@

\paragraph{Partial and Total Parsers}

>>>>

%% \paragraph{Weak and Strong Equivalence}
In the theory of formal grammars, there are two different notions of
equivalence: weak generative capacity, meaning just which strings are
accepted by the grammar, and \emph{strong} generative capacity, when
the parse trees of the two grammars are isomorphic
\cite{chom1963}. Using linear types as grammars, we can define both of
these notions of equivalence in \theoryabbv.
\begin{definition}
  \label{def:weakequiv}
  Grammars $\A$ and $\B$ are \emph{weakly equivalent} if there exists
  parse transformers $\f : \ltonl{(\A \lto \B)}$ and $\g : \ltonl {(\B
    \lto \A)}$. $\A$ is a \emph{retract} of $\B$ if $\A$ and $\B$ are
  weakly equivalent and $\lambda a. g(f(a)) \equiv \lambda a.a$. They
  are \emph{strongly equivalent} if further the other composition is
  the identity, i.e., $\lambda b. f(g(b)) \equiv \lambda b.b$.
\end{definition}

%% That is, grammars are weakly equivalent if we can construct a parse
%% tree of $B$ from a parse tree of $A$ and vice-versa, whereas they are
%% strongly equivalent if these processes are inverse to each other. The
%% definition of retract is a useful intermediate notion. In retrospect,
%% our axiom about $\String$ says that $\String$ is a retract of $\top$.

@@

%% \paragraph{Ambiguity}
A formal grammar $\A$ is ambiguous if there are multiple parse trees
for the \emph{same} string $\w$.  For example, $\a \oplus \a$ is
ambiguous because there are two parses of $\stringquote{a}$,
constructed using $\inl$ and $\inr$. On the other hand, a formal
grammar is unambiguous when there is at most one parse tree for any
input string. We can capture this notion as a type in \theoryabbv in a
clever way:
\begin{definition}
  \label{def:unambig}
  A grammar $\A$ is \emph{unambiguous} if for every linear type $\B$,
  $\f : \ltonl{(\B \lto \A)}$, and $\g : \ltonl{(\B \lto \A)}$
  then $\f \equiv \g$
\end{definition}
\cref{def:unambig} can be read more intuitively as stating that $\A$
is unambiguous if there is at most one way to transform parses of any
other grammar $\B$ into parse of $\A$. This notion of an unambiguous
type is the analog for linear types of the definition of a (homotopy)
\emph{proposition} in the terminology of homotopy type
theory\cite{hottbook}. The most basic unambiguous types are $\top$ and
$0$, and in a system of classical logic all unambiguous types would
have to be equivalent to one of these, but with our axioms we can show
also that $\I$ and literals $\literal c$ are unambiguous. To see this, first, we
establish two useful properties of unambiguity.
\begin{lemma}[\agdalogo]
  If $\B$ is unambiguous and $\A$ is a retract of $\B$ then $\A$ is unambiguous.
  If a disjunction $\bigoplus\limits_{x:X} A(x)$ is unambiguous then
  each $A(x)$ is unambiguous.
\end{lemma}
From the first principle, we can prove that $\String$ is unambiguous,
since it is a retract of $\top$. In fact, observe that if $\A$ is a
retract of $\B$ and $\B$ is unambiguous, then in fact $\A$ and $\B$
are strongly equivalent, as the equation $\lambda b. f(g(b)) \equiv
\lambda b. b$ follows because $\B$ is unambiguous. Therefore
$\String$ is also strongly equivalent to $\top$. Next, since $\String$ is
defined as a Kleene star, we can easily show that $\String \cong \I
\oplus \Char \oplus (\Char \otimes \Char \otimes \String)$. Then by
the second principle we have that $\I$ and $\Char$ and each literal
$\literal c$ are unambiguous as well.

@@

%% \paragraph{Partial and Total Parsers}
